---
title: "Learning R with Chess"
output:
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(chess)
library(rchess)
library(tidyverse)
library(gridExtra)

knitr::opts_chunk$set(echo = TRUE, comment = "")
fen_to_matrix <- function(fen) {
  pieces <- strsplit(fen, ' ')[[1]][1]
  rows   <- strsplit(pieces, "/")[[1]]
  for (i in 1:8) {
    rows <- gsub(i, paste(rep(' ', i), collapse=''), rows)
  }
  matrix(unlist(strsplit(rows, '')), 8, 8, byrow = TRUE, list(c(8:1), letters[1:8]))
}

create_fen_plot <- function(fen, title) {
  mat <- fen_to_matrix(fen)
  unicode <- c(
    ` ` = '',
    P='\u2659', R='\u2656', N='\u2658', B='\u2657', Q='\u2655', K='\u2654', # White
    p='\u265F', r='\u265C', n='\u265E', b='\u265D', q='\u265B', k='\u265A'  # Black
  )
  board <- expand.grid(x=1:8, y=1:8) %>%
    mutate(
      tilecol = (x+y)%%2 == 1,
      piece  = as.vector(t(mat[8:1,])),
      colour = piece %in% letters,
      unicode = unicode[piece]
    )
  ggplot(board, aes(x, y)) + 
    geom_tile(aes(fill = tilecol), colour = 'black', show.legend = FALSE) + 
    geom_text(aes(label = unicode), family="Arial Unicode MS", size = 8) + 
    coord_equal() + 
    theme_void() +
    scale_fill_manual(values = c('#D2B48C', '#F5F5DC'))  +
    labs(title = title) +
    theme(plot.title = element_text(hjust = 0.5))
}

game <- Chess$new()
game$move("e4")
```


## Introduction

Chess is a game that dates back centuries and has captivated people from all walks of life for a long time. It isn't necessary to know the history of the game but if you're interested, here's a quick 5 minute primer.

![](https://www.youtube.com/watch?v=YeB-1F-UKO0)

One reason people become fascinated with chess is that it is an __unsolved__ game. This means there is no perfect way to play.

The game is a mixture of creativity and ruthless logic. It was once thought that the only minds capable of such flexibility would have to be humans. However, in 1997, an IBM machine named _Deep Blue_ defeated Garry Kasparov 3<sup>1</sup>&frasl;<sub>2</sub> - 2<sup>1</sup>&frasl;<sub>2</sub>. This marked the first time a computer had beaten the reigning world champion. Kasparov is still considered (arguably) the greatest chess player ever so this was a turning point for machine intelligence.

So how did _Deep Blue_ do it? That's the subject of this tutorial. Along the way, you'll learn how to do a lot of things in R, including:

* How to interact with objects and environments in R
* How to use lists in R
* How to search through lists
* How to select from a list
* How to use simple conditional statements (eg. _while()_)

## Setting Up The Chessboard

Chess is played on an 8x8 grid of alternately dark and light coloured squares. It is a two-player game in which each player has 16 pieces at the start. Each player takes their turn, starting with the player who has the white pieces. A turn cannot be skipped or passed. These kinds of simple rules are easy for a computer to understand. In fact, the _rchess_ package has these rules coded so we can load up the package and get started!

```{r, echo=TRUE, eval=FALSE}
install.packages("rchess")
library(rchess)
```

To start a chess game we create a Chess environment like this:
```{r, fig.align='center'}
game <- Chess$new()
plot(game)
```

You probably recognise this as a standard chessboard. It is interesting to note that there varieties of chess for 4 players, with different piece configurations and moves. None of those are dealt with here but with the skills you'll learn, maybe you can get that going!

## Making A Move

In Chess, the game always starts with White making a move. An element of the new game we have created in the previous section, called __moves__ simply lists the possible moves in the current game position. We can bring it up like so:

```{r}
game$moves()
```

If we want to find out how many possible moves there are, we can use the __length__ function to find out how many moves are possible. All we need to do is put the list into the length function as follows:

```{r}
length(game$moves())
```

So there are 20 possible moves here. That's a lot to choose from. Let's play the most common opening move in chess and move the King's pawn 2 spaces. This is known in chess notation as _e4_, meaning move the _e_ pawn to the _4^th^_ rank. You may be able to see that is listed as the 10^th^ move in the previous list of 20 possible moves.

```{r}
game$moves()[[10]]
```

We can make a move using the __move__ function within the chess environment like so:

```{r}
game$move("e4")
plot(game)
```

Now pull up the list of possible moves for Black!

```{r find_moves1, exercise = TRUE}

```

If we want to create a computer that can play chess as well as a human being, then we need to give it a way to decide between all these moves.

The function __sample__ will allow us to select a move from this list. If we use sample like this, we're just picking a move at random.

```{r, eval=FALSE}
sample(game$moves(),1)
```

```{r, echo = F}
print("a5")
```

It's now a simple matter of putting all that code to sample the list of moves into the __move__ function just like before.

```{r, eval = F}
game$move(sample(game$moves(),1))
plot(game)
```

```{r, echo = F}
game$move("f6")
plot(game)
```

This time the computer has not played _a5_ because remember, it has _randomly_ selected a move from a list of 20.

## Searching Through Potential Moves

```{r}
g2 <- Chess$new()
g2$move("e4")$move("e5")$
  move("Nf3")$move("Nc6")$
  move("Bc4")$move("d5")
plot(g2)
```

```{r}
g2$moves()
length(g2$moves())
```

```{r}
grep("x", g2$moves())
grep("O-O", g2$moves())
```

```{r}
m <- g2$moves()[c(3,8,10,32)]
m <- g2$moves()[c(grep("x", g2$moves()),grep("O-O", g2$moves()))]
```

```{r, fig.height=8}
g3 <- list()
for(i in 1:length(m)){
  g2$move(m[i])
  g3[i] <- g2$fen()
  g2$undo()
}

grid.arrange(create_fen_plot(g3[[1]],m[1]),
             create_fen_plot(g3[[2]],m[2]),
             create_fen_plot(g3[[3]],m[3]),
             create_fen_plot(g3[[4]],m[4]),
             ncol = 2)
```

## Simulating A Game

```{r}
game <- Chess$new()

while(game$game_over() == FALSE){
 game$move(sample(game$moves(), 1)) 
}

plot(game)


```

```{r}
length(game$history())/2
```

## Simulating Many Games

```{r}
game <- Chess$new()

pgns <- list() 
moves <- list()
  
for(i in 1:2){
  while(game$game_over() == FALSE){
    game$move(sample(game$moves(), 1)) 
  }
  pgns[[i]] <- game$pgn()
  moves[[i]] <- game$history()
  game$reset()
}
```

```{r}
x <- unlist(lapply(moves, length))/2
hist(x)
```




## Challenges

* Create a function that acts as a chess bot
  - selects its own moves
  - allows user input for their moves
  - can play games against other bots
* Evaluate a position and add this capability to the bot
  - get an evaluation (stockfish?)
  - have the bot use evaluation to select moves
  - compare bots of different ability
  - set up tournament of bots
